---
title: "Untitled"
author: "Laura_Hakkinen"
date: "2024-01-10"
output: html_document
---

# RESULTS STEP 11: Test between management the 5 most abundant taxa 

```{r}
library('phyloseq')
library("cowplot")
library("dplyr")
library("ggplot2")
library("vegan")
library("microbiome")
library("tibble")
library(stringr)
library(reshape2)
library(tidyr)

setwd('\\\\ad.helsinki.fi\\home\\l\\lehakkin\\Desktop\\PROJECT_YONI_fungal_ITS')


load('ps_FINAL')
ps

ps_RA <- transform(ps, "compositional")

meta <- meta(ps)
```

# 1. Five most abundant phyla

```{r}

ps_RA_x <- aggregate_rare(ps_RA, level = "phylum", detection = 0, prevalence = 0)
ps_RA_x              
# 14 taxa and 140 samples


#create data table
x_df <-  psmelt(ps_RA_x)

x_df$year <- "2019"

x <- x_df %>%
  group_by(year, OTU) %>%
  summarise(mean = mean(Abundance, na.rm = TRUE), se = (sd(Abundance, na.rm = TRUE)/sqrt(length((Abundance)))))

y <- x %>%
    arrange(desc(mean)) %>%
    slice(1:5)
print(y)
```


## 1.2. Test for each 5 phyla


```{r}
library(car)

taxa <- y$OTU

for (i in taxa) {
 df <- filter(x_df, OTU == i)
 print(i)
 # Using leveneTest()
result = leveneTest(Abundance ~ sample_type, df)
# print the result
print(result)

# perform the Kruskal test
k <- kruskal.test(Abundance ~ sample_type, data = df)
print(k)

w <- pairwise.wilcox.test(df$Abundance, df$sample_type,
                     p.adjust.method = "BH")
print(w)
}

x <- x_df %>%
  filter(OTU %in% taxa) %>% group_by(OTU, sample_type) %>% summarise(mean = mean(Abundance, na.rm = TRUE), se = (sd(Abundance, na.rm = TRUE)/sqrt(length((Abundance)))))
print(x)

```


## 1.3. Test phyla in soil layers

### 1.3.1 meadow, organic and conevntional only

#### 1.3.1.1. Five most abundant phyla

```{r}
x_df_nf <- subset(x_df, sample_type!="forest")

x_df_nf$depth <- as.factor(x_df_nf$depth)
```


```{r}
x_df_nf$year <- "2019"

x <- x_df_nf %>%
  group_by(year, OTU) %>%
  summarise(mean = mean(Abundance, na.rm = TRUE), se = (sd(Abundance, na.rm = TRUE)/sqrt(length((Abundance)))))

y <- x %>%
    arrange(desc(mean)) %>%
    slice(1:5)
print(y)

taxa <- y$OTU

x <- x_df_nf %>%
  filter(OTU %in% taxa) %>% group_by(OTU, depth) %>% summarise(mean = mean(Abundance, na.rm = TRUE), se = (sd(Abundance, na.rm = TRUE)/sqrt(length((Abundance)))))
print(x)

setwd('\\\\ad.helsinki.fi\\home\\l\\lehakkin\\Desktop\\PROJECT_YONI_fungal_ITS')

write.csv2(x, file = "5_Phyla_in_soil_layers_WITHOUT_forest_mean.csv")
```



```{r}
library(car)
library("rcompanion")
library("multcompView")

taxa <- y$OTU

for (i in taxa) {
  df <- subset(x_df_nf, OTU==i)
  # Using leveneTest()
  print(i)
result = leveneTest(Abundance ~ depth, df)
# print the result
print(result)

# perform the Kruskal test
k <- kruskal.test(Abundance ~ depth, data = df)
print(k)

wilcox.res <- pairwise.wilcox.test(df$Abundance, df$depth,
                     p.adjust.method = "BH")
wilcox.res <- wilcox.res[["p.value"]]
print(wilcox.res)
mc = fullPTable(wilcox.res)
mc <- multcompLetters(mc)
mc <- mc[["Letters"]]
print(mc)
}
```


#### 1.3.1.2. AMF below 30 cm

```{r}
taxa <- "Glomeromycota"

# Construct a data.frame with the selected taxonomic group
df <- filter(x_df_nf, OTU == taxa & depth_numerical > 40)

#library(car)
# Using leveneTest()
result = leveneTest(Abundance ~ sample_type, df)
# print the result
print(result)


x <- df %>%
  group_by(sample_type) %>%
  summarise(mean = mean(Abundance, na.rm = TRUE), se = (sd(Abundance, na.rm = TRUE)/sqrt(length((Abundance)))))
print(x)

# one-way ANOVA
res.aov <- aov(Abundance ~ sample_type, data = df)
# Summary of the analysis
print(summary(res.aov))
```


### 1.3.2 Forest only

#### 1.3.2.1. Five most abundant phyla

```{r}
x_df_nf <- subset(x_df, sample_type=="forest")

x_df_nf$depth <- as.factor(x_df_nf$depth)
```


```{r}
x_df_nf$year <- "2019"

x <- x_df_nf %>%
  group_by(year, OTU) %>%
  summarise(mean = mean(Abundance, na.rm = TRUE), se = (sd(Abundance, na.rm = TRUE)/sqrt(length((Abundance)))))

y <- x %>%
    arrange(desc(mean)) %>%
    slice(1:5)
print(y)

taxa <- y$OTU

x <- x_df_nf %>%
  filter(OTU %in% taxa) %>% group_by(OTU, depth) %>% summarise(mean = mean(Abundance, na.rm = TRUE), se = (sd(Abundance, na.rm = TRUE)/sqrt(length((Abundance)))))
print(x)

setwd('\\\\ad.helsinki.fi\\home\\l\\lehakkin\\Desktop\\PROJECT_YONI_fungal_ITS')

write.csv2(x, file = "5_Phyla_in_soil_layers_ONLY_forest_mean.csv")

```



```{r}
taxa <- "Glomeromycota"

# Construct a data.frame with the selected taxonomic group
df <- subset(x_df_nf, OTU==taxa)

#library(car)
# Using leveneTest()
result = leveneTest(Abundance ~ depth, df)
# print the result
print(result)


# perform the Kruskal test
kruskal.test(Abundance ~ depth, data = df)
```

# 2. Class

```{r}

ps_RA_x <- aggregate_rare(ps_RA, level = "class", detection = 0, prevalence = 0)
ps_RA_x              
# 68 taxa and 140 samples


#create data table
x_df <-  psmelt(ps_RA_x)

```


## 2.1. Five most abundant class

```{r}
x_df$year <- "2019"

x <- x_df %>%
  group_by(year, OTU) %>%
  summarise(mean = mean(Abundance, na.rm = TRUE), se = (sd(Abundance, na.rm = TRUE)/sqrt(length((Abundance)))))

y <- x %>%
    arrange(desc(mean)) %>%
    slice(1:5)
print(y)

taxa <- y$OTU

x <- x_df %>%
  filter(OTU %in% taxa) %>% group_by(OTU, sample_type) %>% summarise(mean = mean(Abundance, na.rm = TRUE), se = (sd(Abundance, na.rm = TRUE)/sqrt(length((Abundance)))))
print(x)

setwd('\\\\ad.helsinki.fi\\home\\l\\lehakkin\\Desktop\\PROJECT_YONI_fungal_ITS')

write.csv2(x, file = "5_Class_in_soil_layers_All_Management.csv")
```

## 2.2. Test for each class separately

```{r}

taxa <- y$OTU

for (i in taxa) {
  # Construct a data.frame with the selected taxonomic group
df <- filter(x_df, OTU == i)
print(i)

# Using leveneTest()
result = leveneTest(Abundance ~ sample_type, df)
# print the result
print(result)


# perform the Kruskal test
k <- kruskal.test(Abundance ~ sample_type, data = df)
print(k)

w <- pairwise.wilcox.test(df$Abundance, df$sample_type,
                     p.adjust.method = "BH")
print(w)
}

x <- x_df %>%
  filter(OTU %in% taxa) %>% group_by(OTU, sample_type) %>% summarise(mean = mean(Abundance, na.rm = TRUE), se = (sd(Abundance, na.rm = TRUE)/sqrt(length((Abundance)))))
print(x)
```



## 2.3. Test classes in soil layers

### 2.3.1 meadow, organic and conevntional only

#### 2.3.1.1. Five most abundant classes

```{r}
x_df_nf <- subset(x_df, sample_type!="forest")

x_df_nf$depth <- as.factor(x_df_nf$depth)
```


```{r}
x_df_nf$year <- "2019"

x <- x_df_nf %>%
  group_by(year, OTU) %>%
  summarise(mean = mean(Abundance, na.rm = TRUE), se = (sd(Abundance, na.rm = TRUE)/sqrt(length((Abundance)))))

y <- x %>%
    arrange(desc(mean)) %>%
    slice(1:5)
print(y)

taxa <- y$OTU

x <- x_df_nf %>%
  filter(OTU %in% taxa) %>% group_by(OTU, depth) %>% summarise(mean = mean(Abundance, na.rm = TRUE), se = (sd(Abundance, na.rm = TRUE)/sqrt(length((Abundance)))))
print(x)

setwd('\\\\ad.helsinki.fi\\home\\l\\lehakkin\\Desktop\\PROJECT_YONI_fungal_ITS')

write.csv2(x, file = "5_Classes_in_soil_layers_WITHOUT_forest_mean.csv")

```



```{r}
taxa <- y$OTU

for (i in taxa) {
  # Construct a data.frame with the selected taxonomic group
df <- subset(x_df_nf, OTU==i)
print(i)

# Using leveneTest()
result = leveneTest(Abundance ~ depth, df)
# print the result
print(result)

# perform the Kruskal test
k <- kruskal.test(Abundance ~ depth, data = df)
print(k)

wilcox.res <- pairwise.wilcox.test(df$Abundance, df$depth,
                     p.adjust.method = "BH")
wilcox.res <- wilcox.res[["p.value"]]
print(wilcox.res)

mc = fullPTable(wilcox.res)
mc <- multcompLetters(mc)
mc <- mc[["Letters"]]
print(mc)
}

```



### 2.3.2 Forest only

#### 2.3.2.1. Five most abundant classes

```{r}
x_df_nf <- subset(x_df, sample_type=="forest")

x_df_nf$depth <- as.factor(x_df_nf$depth)
```


```{r}
x_df_nf$year <- "2019"

x <- x_df_nf %>%
  group_by(year, OTU) %>%
  summarise(mean = mean(Abundance, na.rm = TRUE), se = (sd(Abundance, na.rm = TRUE)/sqrt(length((Abundance)))))

y <- x %>%
    arrange(desc(mean)) %>%
    slice(1:5)
print(y)

taxa <- y$OTU

x <- x_df_nf %>%
  filter(OTU %in% taxa) %>% group_by(OTU, depth) %>% summarise(mean = mean(Abundance, na.rm = TRUE), se = (sd(Abundance, na.rm = TRUE)/sqrt(length((Abundance)))))
print(x)

setwd('\\\\ad.helsinki.fi\\home\\l\\lehakkin\\Desktop\\PROJECT_YONI_fungal_ITS')

write.csv2(x, file = "5_Classes_in_soil_layers_ONLY_forest_mean.csv")
```




```{r}

taxa <- y$OTU

for (i in taxa) {
  # Construct a data.frame with the selected taxonomic group
df <- subset(x_df_nf, OTU==i)
print(i)

# Using leveneTest()
result = leveneTest(Abundance ~ depth, df)
# print the result
print(result)

# perform the Kruskal test
k <- kruskal.test(Abundance ~ depth, data = df)
print(k)

wilcox.res <- pairwise.wilcox.test(df$Abundance, df$depth,
                     p.adjust.method = "BH")
wilcox.res <- wilcox.res[["p.value"]]
print(wilcox.res)

mc = fullPTable(wilcox.res)
mc <- multcompLetters(mc)
mc <- mc[["Letters"]]
print(mc)
}

```


### 2.3.3 Test Classes that popped up in certain layers

```{r}

ps_RA_x <- aggregate_rare(ps_RA, level = "class", detection = 0, prevalence = 0)
ps_RA_x              
# 68 taxa and 140 samples


#create data table
x_df <-  psmelt(ps_RA_x)

```
This senetence based on looking at the class composition barplot:

"Fungal classes that had mostly low proportion but popped up in specific layers were Pezizomycetes in forest 30-40 cm layer (ANOVA; p = 0.006), Archaeosporomycetes in organic 30-80 cm, Geoglossomycetes in forest 10-20 cm, meadow 10-40 cm and organic 20-80 cm, Microbotryomycetes in conventional 40-80 cm, and Glomeromycetes in meadow 30-80 cm, and Orbiliomycetes."

Let's test them separately

#### 2.3.3.1. Pezizomycetes

```{r}
x_df_nf <- subset(x_df, sample_type=="forest")

x_df_nf$compare <- NA

x_df_nf$compare[x_df_nf$depth=="30...40"]<-"yes"
x_df_nf$compare[is.na(x_df_nf$compare)]<-"no"
```


```{r}
taxa <- "Pezizomycetes"

# Construct a data.frame with the selected taxonomic group
df <- subset(x_df_nf, OTU==taxa)

#library(car)
# Using leveneTest()
result = leveneTest(Abundance ~ compare, df)
# print the result
print(result)

# first one-way ANOVA
res.aov <- aov(Abundance ~ compare, data = df)
# Summary of the analysis
print(summary(res.aov))

```

This sentence remains to be tested:

"Fungal classes that had mostly low proportion but popped up in specific layers were Pezizomycetes in forest 30-40 cm layer (ANOVA; p = 0.006), Archaeosporomycetes in organic 30-80 cm, Geoglossomycetes in forest 10-20 cm, meadow 10-40 cm and organic 20-80 cm, Microbotryomycetes in conventional 40-80 cm, and Glomeromycetes in meadow 30-80 cm, and Orbiliomycetes."


#### 2.3.3.2. Archaeosporomycetes

```{r}
x_df_nf <- subset(x_df, sample_type=="organic")

x_df_nf$compare <- NA

x_df_nf$compare[x_df_nf$depth=="30...40"]<-"yes"
x_df_nf$compare[is.na(x_df_nf$compare)]<-"no"
```


```{r}
taxa <- "Archaeosporomycetes"

# Construct a data.frame with the selected taxonomic group
df <- subset(x_df_nf, OTU==taxa)

#library(car)
# Using leveneTest()
result = leveneTest(Abundance ~ compare, df)
# print the result
print(result)

# perform the Kruskal test
k <- kruskal.test(Abundance ~ compare, data = df)
print(k)
```

Not significant!

This sentence remains to be tested:

"Fungal classes that had mostly low proportion but popped up in specific layers were Pezizomycetes in forest 30-40 cm layer (ANOVA; p = 0.006), Geoglossomycetes in meadow 10-40 cm, Microbotryomycetes in conventional 40-80 cm, and Glomeromycetes in meadow 30-80 cm, and Orbiliomycetes."


#### 2.3.3.3. Geoglossomycetes

not sig. in:

- forest 10-20 cm
- organic 20-80 cm

```{r}
x_df_nf <- subset(x_df, sample_type=="meadow")

x_df_nf$compare <- NA

x_df_nf$compare[x_df_nf$depth=="20...30"]<-"yes"
x_df_nf$compare[x_df_nf$depth=="30...40"]<-"yes"
x_df_nf$compare[x_df_nf$depth=="10...20"]<-"yes"
x_df_nf$compare[is.na(x_df_nf$compare)]<-"no"
```


```{r}
taxa <- "Geoglossomycetes"

# Construct a data.frame with the selected taxonomic group
df <- subset(x_df_nf, OTU==taxa)

#library(car)
# Using leveneTest()
result = leveneTest(Abundance ~ compare, df)
# print the result
print(result)

# perform the Kruskal test
k <- kruskal.test(Abundance ~ compare, data = df)
print(k)
```

Is sig. in meadow 10-40 cm


"Fungal classes that had mostly low proportion but popped up in specific layers were Pezizomycetes in forest 30-40 cm layer (ANOVA; p = 0.006), Geoglossomycetes meadow 10-40 cm (Kruskal; p = 0.000), Microbotryomycetes in conventional 40-80 cm (Kruskal, P = 0.030), and Glomeromycetes in meadow 30-80 cm, and Orbiliomycetes."


#### 2.3.3.4. Microbotryomycetes

```{r}
x_df_nf <- subset(x_df, sample_type=="conventional")

x_df_nf$compare <- NA

x_df_nf$compare[x_df_nf$depth=="40..."]<-"yes"
x_df_nf$compare[is.na(x_df_nf$compare)]<-"no"
```


```{r}
taxa <- "Microbotryomycetes"

# Construct a data.frame with the selected taxonomic group
df <- subset(x_df_nf, OTU==taxa)

#library(car)
# Using leveneTest()
result = leveneTest(Abundance ~ compare, df)
# print the result
print(result)

# perform the Kruskal test
k <- kruskal.test(Abundance ~ compare, data = df)
print(k)
```

Fungal classes that had mostly low proportion but popped up in specific layers were Pezizomycetes in forest 30-40 cm layer (ANOVA; p = 0.006), Geoglossomycetes meadow 10-40 cm (Kruskal; p = 0.000), Microbotryomycetes in conventional 40-80 cm (Kruskal, P = 0.030), and Glomeromycetes in meadow 30-80 cm, and Orbiliomycetes.


#### 2.3.3.5. Glomeromycetes

```{r}
x_df_nf <- subset(x_df, sample_type=="meadow")

x_df_nf$compare <- NA

x_df_nf$compare[x_df_nf$depth=="30...40"]<-"yes"
x_df_nf$compare[x_df_nf$depth=="40..."]<-"yes"
x_df_nf$compare[is.na(x_df_nf$compare)]<-"no"
```


```{r}
taxa <- "Glomeromycetes"

# Construct a data.frame with the selected taxonomic group
df <- subset(x_df_nf, OTU==taxa)

#library(car)
# Using leveneTest()
result = leveneTest(Abundance ~ compare, df)
# print the result
print(result)

# perform the Kruskal test
k <- kruskal.test(Abundance ~ compare, data = df)
print(k)
```
Not sig.

Fungal classes that had mostly low proportion but popped up in specific layers were Pezizomycetes in forest 30-40 cm layer (ANOVA; p = 0.006), Geoglossomycetes meadow 10-40 cm (Kruskal; p = 0.000), Microbotryomycetes in conventional 40-80 cm (Kruskal, P = 0.030), and Orbiliomycetes in meadow 10-30 cm.


#### 2.3.3.6. Orbiliomycetes.

```{r}
x_df_nf <- subset(x_df, sample_type=="meadow")

x_df_nf$compare <- NA

x_df_nf$compare[x_df_nf$depth=="10...20"]<-"yes"
x_df_nf$compare[x_df_nf$depth=="20...30"]<-"yes"
x_df_nf$compare[is.na(x_df_nf$compare)]<-"no"
```


```{r}
taxa <- "Orbiliomycetes"

# Construct a data.frame with the selected taxonomic group
df <- subset(x_df_nf, OTU==taxa)

#library(car)
# Using leveneTest()
result = leveneTest(Abundance ~ compare, df)
# print the result
print(result)

# perform the Kruskal test
k <- kruskal.test(Abundance ~ compare, data = df)
print(k)
```


This sentence remains valid:

"Fungal classes that had mostly low proportion but popped up in specific layers were Pezizomycetes in forest 30-40 cm layer (ANOVA; p = 0.006), Geoglossomycetes meadow 10-40 cm (Kruskal; p = 0.000), Microbotryomycetes in conventional 40-80 cm (Kruskal, P = 0.030), and Orbiliomycetes in meadow 10-30 cm (Kruskal; p = 0.000)."




