---
title: "Untitled"
author: "Laura_Hakkinen"
date: "2024-06-18"
output: html_document
---

# RESULTS STEP 13: AMF bubble plot

# 1. Load data and add soil_type_depth

```{r}
library(vegan)
library(goeveg)
library(metagMisc)
library(phyloseq)
library('phyloseq')
library("cowplot")
library("dplyr")
library("ggplot2")
library("vegan")
library("microbiome")
library("tibble")
library(stringr)
library(reshape2)
library(tidyr)
library(car)

setwd('\\\\ad.helsinki.fi\\home\\l\\lehakkin\\Desktop\\PROJECT_YONI_fungal_ITS')

load('ps_FINAL')
ps

# add sample_type_depth# add soil_type_depth
sample_data(ps)$soil_type_depth <- paste(sample_data(ps)$sample_type, "_", sample_data(ps)$depth)

meta <- meta(ps)
```


# 2. Keep only Glomeromycota

Note! *first* transform to RA and *then* filter AMF


```{r}
# Relative abundance
ps_RA <- microbiome::transform(ps, 'compositional')
ps_RA

# keep only Glomeromycota
ps2_std_G <- subset_taxa(ps_RA, phylum=="Glomeromycota")
ps2_std_G
# 263 taxa and 140 samples
```


# 3. Aggregate samples by soil_type_depth

For the bubble plot, we want to get average values based on soil_type_depth

```{r}
library("metagMisc")

ps2_std_G <- phyloseq_average(
  ps2_std_G,
  avg_type = "arithmetic",
  group = "soil_type_depth",
  drop_group_zero = FALSE,
  verbose = FALSE,
  progress = NULL
)

ps2_std_G
```


```{r}
# aggregate

ps2_std_G <- aggregate_rare(ps2_std_G, level = 'genus', detection = 0, prevalence = 0, include.lowest = TRUE)
ps2_std_G
```


# 4. Extract tax and otu tables

```{r}
tax_table_G <- as.data.frame(as.matrix(tax_table(ps2_std_G)))
OTU_genus_table_G <- as.data.frame(as.matrix(otu_table(ps2_std_G)))
```


# 5. Add total AMF to otu table and make wide

```{r}
# make OTUs into columns
OTU_genus_table_G <- t(OTU_genus_table_G)
OTU_genus_table_G <- as.data.frame(OTU_genus_table_G)

# add total of phylum glomeracea
OTU_genus_table_G <- cbind(
  OTU_genus_table_G,
  total = rowSums(OTU_genus_table_G
))

# Change your data structure from a wide format to a long format. Put any variables that are not OTUs/genus, into the id parameter

#convert data frame from a wide format to a long format
pc <- tibble::rownames_to_column(OTU_genus_table_G, "Sample")
pcm = melt(pc, id = c("Sample"))
```


# 6. Combine otu wide and meta

```{r}
meta_x <- meta[!duplicated(meta$soil_type_depth),]
# Vector of column names to select
columns_to_keep <- c("depth", "sample_type", "soil_type_depth")

# Using base R to select columns
meta_x <- meta_x[, columns_to_keep]

rownames(meta_x) <- NULL

# add metadata
pcm <- left_join(pcm, meta_x, by=c("Sample"="soil_type_depth"))

# change the column name
colnames(pcm)[1] <- "soil_type_depth"

```



# 7. Reorder soil_type_depth

```{r}
pcm$soil_type_depth <- as.factor(pcm$soil_type_depth)

# Change the order of samples
pcm$soil_type_depth <- factor(pcm$soil_type_depth,levels=c("forest _ 0...10", "forest _ 10...20", "forest _ 20...30", "forest _ 30...40", "forest _ 40...", "meadow _ 0...10", "meadow _ 10...20", "meadow _ 20...30", "meadow _ 30...40", "meadow _ 40...", "organic _ 0...10", "organic _ 10...20", "organic _ 20...30", "organic _ 30...40", "organic _ 40...", "conventional _ 0...10", "conventional _ 10...20", "conventional _ 20...30", "conventional _ 30...40", "conventional _ 40..."))

levels(pcm$soil_type_depth)
```


# 8. Add other taxonomic levels

```{r}
# add the other taxonomic level annotations
TAX <- as.data.frame(as.matrix(tax_table(ps)))
row.names(TAX) <- NULL
TAX <- TAX[, 1:6]
TAX <- filter(TAX, phylum == "Glomeromycota")


TAX <- TAX[!duplicated(TAX$genus),]
pcm2 <- left_join(pcm, TAX, by = c("variable" = "genus"))
```



# 9. Add the taxonomic rank letter to genus

```{r}
pcm2$variable[pcm2$variable == "Ambisporaceae_unclassified"] <- "Ambisporaceae (f)"
pcm2$variable[pcm2$variable == "Archaeospora"] <- "Archaeospora (g)"
pcm2$variable[pcm2$variable == "Archaeosporaceae_unclassified"] <- "Archaeosporaceae (f)"
pcm2$variable[pcm2$variable == "Archaeosporales_unclassified"] <- "Archaeosporales (o)"
pcm2$variable[pcm2$variable == "Diversispora"] <- "Diversispora (g)"
pcm2$variable[pcm2$variable == "Claroideoglomus"] <- "Claroideoglomus (g)"
pcm2$variable[pcm2$variable == "Entrophospora"] <- "Entrophospora (g)"
pcm2$variable[pcm2$variable == "Claroideoglomeraceae_unclassified"] <- "Claroideoglomeraceae (f)"
pcm2$variable[pcm2$variable == "Dominikia"] <- "Dominikia (g)"
pcm2$variable[pcm2$variable == "Funneliformis"] <- "Funneliformis (g)"
pcm2$variable[pcm2$variable == "Glomeraceae_unclassified"] <- "Glomeraceae (f)"
pcm2$variable[pcm2$variable == "Glomus"] <- "Glomus (g)"
pcm2$variable[pcm2$variable == "Microdominikia"] <- "Microdominikia (g)"
pcm2$variable[pcm2$variable == "Rhizophagus"] <- "Rhizophagus (g)"
pcm2$variable[pcm2$variable == "Glomeromycota_unclassified"] <- "Glomeromycota (p)"
pcm2$variable[pcm2$variable == "Paraglomus"] <- "Paraglomus (g)"
pcm2$variable[pcm2$variable == "Paraglomerales_unclassified"] <- "Paraglomerales (o)"
pcm2$variable[pcm2$variable == "total"] <- "total"
```


```{r}
# add total to family
pcm2$family <- pcm2$family %>% replace_na('total')

```


# 10. Change family names

```{r}
pcm2$family[pcm2$family=="Ambisporaceae"] <- "Ambisporaceae (f)"
pcm2$family[pcm2$family=="Archaeosporaceae"] <- "Archaeosporaceae (f)"
pcm2$family[pcm2$family=="Archaeosporales_unclassified"] <- "Archaeosporales (o)"
pcm2$family[pcm2$family=="Diversisporaceae"] <- "Diversisporaceae (f)"
pcm2$family[pcm2$family=="Entrophosporaceae"] <- "Entrophosporaceae (f)"
pcm2$family[pcm2$family=="Claroideoglomeraceae"] <- "Claroideoglomeraceae (f)"
pcm2$family[pcm2$family=="Glomeraceae"] <- "Glomeraceae (f)"
pcm2$family[pcm2$family=="Glomeromycota_unclassified"] <- "Glomeromycota (p)"
pcm2$family[pcm2$family=="Paraglomeraceae"] <- "Paraglomeraceae (f)"
pcm2$family[pcm2$family=="Paraglomerales_unclassified"] <- "Paraglomerales (o)"
```


```{r}
# make into factor
pcm2$family <- as.factor(pcm2$family)

levels(pcm2$family)
```

# 11. Change Family order

```{r}
# Change level family

pcm2$family <- factor(pcm2$family, levels = c("Ambisporaceae (f)", "Archaeosporaceae (f)", "Diversisporaceae (f)", "Glomeraceae (f)", "Claroideoglomeraceae (f)", "Entrophosporaceae (f)", "Paraglomeraceae (f)", "Glomeromycota (p)", "Archaeosporales (o)", "Paraglomerales (o)", "total"))
```


# 12. Change genus order

```{r}
# make variable into factor
pcm2$variable <- as.factor(pcm2$variable)

```


```{r}
# Change genus level order

pcm2$variable <- factor(pcm2$variable, levels = c("Archaeosporales (o)", "Ambisporaceae (f)", "Archaeosporaceae (f)", "Archaeospora (g)", "Diversispora (g)", "Glomeraceae (f)", "Dominikia (g)", "Funneliformis (g)", "Glomus (g)", "Microdominikia (g)", "Rhizophagus (g)", "Claroideoglomeraceae (f)", "Claroideoglomus (g)", "Entrophospora (g)", "Paraglomerales (o)", "Paraglomus (g)", "Glomeromycota (p)", "total"))

levels(pcm2$variable)
```


# 13. Assign AMF guild based on AMF family

```{r}
pcm2$AMF_guild <- NA

pcm2$AMF_guild[pcm2$family=="Ambisporaceae (f)"] <- "ancestral"
pcm2$AMF_guild[pcm2$family=="Archaeosporaceae (f)"] <- "ancestral"
pcm2$AMF_guild[pcm2$family=="Archaeosporales (o)"] <- "unknown"
pcm2$AMF_guild[pcm2$family=="Diversisporaceae (f)"] <- "edaphophilic"
pcm2$AMF_guild[pcm2$family=="Claroideoglomeraceae (f)"] <- "rhizophilic"
pcm2$AMF_guild[pcm2$family=="Entrophosporaceae (f)"] <- "rhizophilic"
pcm2$AMF_guild[pcm2$family=="Glomeraceae (f)"] <- "rhizophilic"
pcm2$AMF_guild[pcm2$family=="Glomeromycota (p)"] <- "unknown"
pcm2$AMF_guild[pcm2$family=="Paraglomeraceae (f)"] <- "rhizophilic"
pcm2$AMF_guild[pcm2$family=="Paraglomerales (o)"] <- "unknown"


```


# 14. Add new depth

```{r}
pcm2$new_depth <- NA
pcm2$new_depth[pcm2$depth=="0...10"]<-"0-10"
pcm2$new_depth[pcm2$depth=="10...20"]<-"10-20"
pcm2$new_depth[pcm2$depth=="20...30"]<-"20-30"
pcm2$new_depth[pcm2$depth=="30...40"]<-"30-40"
pcm2$new_depth[pcm2$depth=="40..."]<-"40-80"
```


# 15. Modify tax level: Order

Currently, AMF are divided in five orders (Archaeosporales, Diversisporales, Glomerales, Entrophosporales and Paraglomerales)

```{r}
# modifying the column of data frame
pcm2$order <- as.factor(pcm2$order)

levels(pcm2$order)

# Change levels
pcm2$order <- factor(pcm2$order, levels = c("Archaeosporales", "Diversisporales", "Entrophosporales", "Glomerales", "Paraglomerales", "Glomeromycota_unclassified"), labels = c("Archaeosporales (o)", "Diversisporales (o)", "Entrophosporales (o)", "Glomerales (o)", "Paraglomerales (o)", "Glomeromycota (p)"))

# add order = Glomeromycota to total
pcm2$order <- pcm2$order %>% replace_na('Glomeromycota (p)')

levels(pcm2$order)
```


# 16. Plot the Bubble plot

For a bubble plot, you are using geom_point and scaling the size to your value (relative abundance) column.

I checked that:
- thesmallest non-zero is 8.314653e-07 or 0.0000008314653
- and the biggest value is 1.164825e-01 or 0.1164825

So lets set the limits in the figure accordingly

```{r fig.dim=c(14,10)}

# color for AMF_guild
MyPalette = c("red",  "blue", "#ff028d", "black")

# where ("Ancestral" = "red", "Rhizophilic" = "#ff028d", "Edaphophilic" =  "blue", "Unknown" = "black")

xx2 = ggplot(pcm2, aes(x = new_depth, y = variable)) + 
  geom_point(aes(size = value, fill = order, color = AMF_guild), shape = 21, alpha = 1, stroke = 0) + 
  scale_fill_manual(values=c("#99CC99", "#83adb5", "#692D6B", "#D094D2", "#cba69e", "#A9A9A9")) + 
  scale_size_continuous(limits = c(0.0000001, 0.2), range = c(3,30), breaks = c(0.0000001, 0.0001, 0.001, 0.01, 0.2)) + 
  labs( x= "depth (cm)", y = "", size = "Relative abundance", fill = "", color = "AMF guild")  + 
  theme(legend.key=element_blank(), 
        axis.text.x = element_text(colour = "black", size = 14, angle = 45, vjust = 1, hjust = 1),
        axis.title.x=element_text(colour = "black", size = 14,face="bold"), 
        axis.text.y = element_text(colour = "black", size = 14), 
        legend.text = element_text(size = 14, colour ="black"), 
        legend.title = element_text(size = 16, face = "bold"), 
        panel.background = element_blank(), panel.border = element_rect(colour = "black", fill = NA, size = 1.2)) + facet_wrap(vars(sample_type), nrow = 1, ncol = 4) +
  scale_y_discrete(limits = rev(levels(pcm2$variable))) + guides(fill = guide_legend(order = 1, ncol = 2, override.aes = list(size = 8, shape = 21)), color = guide_legend(order = 2, ncol = 2, override.aes = list(shape = 15, size = 8, stroke = 2)))
#+ theme(legend.margin=margin(1,1,1,1), legend.box.spacing = unit(1.6, "pt"))
#+ theme(legend.title.align=0.5)#+ theme(legend.position="bottom")
f1 <- xx2 + scale_color_manual(values = (MyPalette)) + theme(strip.text = element_text(size = 16, color = "black"))
#+  guides(color = guide_legend(nrow = 2, override.aes = list(shape = 15, size = 2, stroke = 2)))   #+ coord_flip() + guide_legend(ncol=2)

### change y axis label colors based on AMF guild

f2 <- f1 + theme(axis.text.y = element_text(color = c("black", "black", "#ff028d", "black", "#ff028d", "#ff028d", "#ff028d", "#ff028d", "#ff028d", "#ff028d", "#ff028d", "#ff028d", "#ff028d", "blue", "red", "red", "red", "black")))

print(f2)
```

Save with 1400 width and 550 height



# 17. Statistical testing

I have previously tested the AMF genera, family, order and guild, but no other difference was found other than:

More Ambisporaceae in forest compared to other Wilcoxon).

So I will not include the tests here, They were done similarly as before e.g. for AMF relative abundance.





