---
title: "Untitled"
author: "Laura_Hakkinen"
date: "2024-01-18"
output: html_document
---

# RESULTS STEP 4: PERMANOVA analysis with soil properties

PERMANOVA with soil properties will be done with only meadow, organic and conventional management types excluding forest

## 1. load packages and data
```{r}

library('phyloseq')
library("cowplot")
library("dplyr")
library("ggplot2")
library("vegan")
library("microbiome")
library("tibble")
library("pairwiseAdonis")

setwd('\\\\ad.helsinki.fi\\home\\l\\lehakkin\\Desktop\\PROJECT_YONI_fungal_ITS')

load('ps_FINAL')
ps


# Pick relative abundances (compositional) and sample metadata 
ps_RA <- microbiome::transform(ps, "compositional")
ps_RA_nf <- subset_samples(ps_RA, sample_type!="forest")

# You have only submitted samples, not OTUs. Define prevalence of each taxa
# (in how many samples did each taxa appear at least once)
prev0 = apply(X = otu_table(ps_RA_nf),
              MARGIN = ifelse(taxa_are_rows(ps_RA_nf), yes = 1, no = 2),
              FUN = function(x){sum(x > 0)})

# Execute prevalence filter, using `prune_taxa()` function
ps_RA_nf = prune_taxa((prev0 > 0), ps_RA_nf)
ps_RA_nf

otu <- abundances(ps_RA_nf)
meta <- meta(ps_RA_nf)
```


**Note:** adonis cannot handle or account for NA or blanks in your data. Use na.omit(meta) %>% before each run where such variables are used where NAs exist


## 2. Calculate Bray-Curtis (dis)similarities

```{r}
ps_RA_bray <- phyloseq::distance(ps_RA_nf, method = "bray")
```


### 4.2 PERMANOVA with soil properties

I will use the following soil properties
                           
"log_root"                                      
"pH_H2O"                                        
"C_g_per_kg"                                    
"N_gkg"                                         
"TP_gkg"                                        
"Alox_mmolkg"                                   
"Feox_mmolkg"                                   
"PH2O_mgkg"                                     
"Porg_mgkg"                                     
"DOC_mgkg"                                      
"Pinorg_mgkg"                                   
"C_per_N"


```{r}
final <- adonis2(formula = ps_RA_bray ~ log_root, data = meta, permutations = 9999, method = "bray")
final
```


```{r}
final <- adonis2(formula = ps_RA_bray ~ pH_H2O, data = meta, permutations = 9999, method = "bray")
final
```


```{r}
final <- adonis2(formula = ps_RA_bray ~ C_g_per_kg, data = meta, permutations = 9999, method = "bray")
final
```


```{r}
final <- adonis2(formula = ps_RA_bray ~ N_gkg, data = meta, permutations = 9999, method = "bray")
final
```


```{r}
final <- adonis2(formula = ps_RA_bray ~ TP_gkg, data = meta, permutations = 9999, method = "bray")
final
```


```{r}
final <- adonis2(formula = ps_RA_bray ~ Alox_mmolkg, data = meta, permutations = 9999, method = "bray")
final
```


```{r}
final <- adonis2(formula = ps_RA_bray ~ Feox_mmolkg, data = meta, permutations = 9999, method = "bray")
final
```

```{r}
final <- adonis2(formula = ps_RA_bray ~ DOC_mgkg, data = meta, permutations = 9999, method = "bray")
final
```

```{r}
final <- adonis2(formula = ps_RA_bray ~ Pinorg_mgkg, data = meta, permutations = 9999, method = "bray")
final
```

```{r}
final <- adonis2(formula = ps_RA_bray ~ C_per_N, data = meta, permutations = 9999, method = "bray")
final
```
                                      

# if missing values

```{r}
colSums(is.na(meta))
```

These are NA:

Porg_mgkg for sample NG2A2_30to40

PH2O_mgkg for sample NG2B3_0to10

```{r}


m <- meta %>% drop_na(Porg_mgkg)

x <- subset_samples(ps_RA_nf, sampleID!="NG2A2_30to40")

# You have only submitted samples, not OTUs. Define prevalence of each taxa
# (in how many samples did each taxa appear at least once)
prev0 = apply(X = otu_table(x),
              MARGIN = ifelse(taxa_are_rows(x), yes = 1, no = 2),
              FUN = function(x){sum(x > 0)})

# Execute prevalence filter, using `prune_taxa()` function
x = prune_taxa((prev0 > 0), x)
x

otu <- abundances(x)

b <- phyloseq::distance(x, method = "bray")

final <- adonis2(formula = b ~ Porg_mgkg, data = m, permutations = 9999, method = "bray", by = "terms")
final
```


```{r}
m <- meta %>% drop_na(PH2O_mgkg)

x <- subset_samples(ps_RA_nf, sampleID!="NG2B3_0to10")

# You have only submitted samples, not OTUs. Define prevalence of each taxa
# (in how many samples did each taxa appear at least once)
prev0 = apply(X = otu_table(x),
              MARGIN = ifelse(taxa_are_rows(x), yes = 1, no = 2),
              FUN = function(x){sum(x > 0)})

# Execute prevalence filter, using `prune_taxa()` function
x = prune_taxa((prev0 > 0), x)
x

otu <- abundances(x)

b <- phyloseq::distance(x, method = "bray")

final <- adonis2(formula = b ~ PH2O_mgkg, data = m, permutations = 9999, method = "bray", by = "terms")
final
```


# PERMANOVA with all soil properties in one model

remove these samples

Porg_mgkg (NG2A2_30to40)
PH2O_mgkg (NG2B3_0to10)


```{r}


m <- meta %>% drop_na(PH2O_mgkg)
m <- m %>% drop_na(Porg_mgkg)

ps_x <- subset_samples(ps_RA_nf, sampleID!="NG2B3_0to10")
ps_x <- subset_samples(ps_x, sampleID!="NG2A2_30to40")

# You have only submitted samples, not OTUs. Define prevalence of each taxa
# (in how many samples did each taxa appear at least once)
prev0 = apply(X = otu_table(ps_x),
              MARGIN = ifelse(taxa_are_rows(ps_x), yes = 1, no = 2),
              FUN = function(x){sum(x > 0)})

# Execute prevalence filter, using `prune_taxa()` function
ps_x = prune_taxa((prev0 > 0), ps_x)
ps_x

b <- phyloseq::distance(ps_x, method = "bray")

final <- adonis2(formula = b ~ log_root + pH_H2O + C_g_per_kg + N_gkg + TP_gkg + Alox_mmolkg + Feox_mmolkg + PH2O_mgkg + Porg_mgkg + DOC_mgkg + Pinorg_mgkg + C_per_N, data = m, permutations = 9999, method = "bray", by = NULL)

final
```

# PERMANOVA for each soil layer separatley

run for all layers and all of these separately: 

```{r}

# Define the environmental variables as a character vector, not as a factor
env <- c("log_root", "pH_H2O", "C_g_per_kg", "N_gkg", "TP_gkg", "Alox_mmolkg", "Feox_mmolkg", 
         "PH2O_mgkg", "Porg_mgkg", "DOC_mgkg", "Pinorg_mgkg", "C_per_N")

# Convert the 'depth' column to a factor
meta$depth <- as.factor(meta$depth)

# Initialize an empty list to store the results
adonis_results <- list()

# Loop over each depth and environmental variable
for (i in levels(meta$depth)) {
  for (j in env) {
    # Subset samples
    ps_x <- subset_samples(ps_RA, sample_type != "forest")
    ps_x <- subset_samples(ps_x, sampleID != "NG2B3_0to10")
    ps_x <- subset_samples(ps_x, sampleID != "NG2A2_30to40")
    ps_x <- subset_samples(ps_x, depth == i)
    meta_subset <- meta(ps_x)

    # Define prevalence of each taxa (in how many samples did each taxa appear at least once)
    prev0 <- apply(X = otu_table(ps_x),
                   MARGIN = ifelse(taxa_are_rows(ps_x), yes = 1, no = 2),
                   FUN = function(x) { sum(x > 0) })

    # Execute prevalence filter, using `prune_taxa()` function
    ps_x <- prune_taxa((prev0 > 0), ps_x)

    # Calculate Bray-Curtis distance
    b <- phyloseq::distance(ps_x, method = "bray")

    # Run adonis2 analysis
    formula <- as.formula(paste("b ~", j))
    adonis_result <- adonis2(formula = formula, data = meta_subset, permutations = 9999, method = "bray", by = NULL)
    
    # Store the result in the list with a descriptive name
    result_name <- paste("depth", i, "env", j, sep = "_")
    adonis_results[[result_name]] <- adonis_result
  }
}

# View the list of results
adonis_results

```



