---
title: "Untitled"
author: "Laura_Hakkinen"
date: "2024-05-27"
output: html_document
---

# RESULTS STEP 1: VENN diagrams and number of OTUs

Here, I will check the number of reads and OTUs and construct venn-diagrams for sample types (or management type) and soil layers. 

```{r}
library("ggpubr")
library('phyloseq')
library("cowplot")
library("dplyr")
library("ggplot2")
library("vegan")
library("microbiome")
library("tibble")
library(stringr)
library(reshape2)
library(tidyr)
library("MicEco")

setwd('\\\\ad.helsinki.fi\\home\\l\\lehakkin\\Desktop\\PROJECT_YONI_fungal_ITS')

load('ps_FINAL')
ps

ps_RA <- microbiome::transform(ps, "compositional")

meta <- meta(ps)
```


# 1. How many reads and OTUs?

```{r}
summarize_phyloseq(ps)
```

We obtained 11662127 fungal reads which clustered into 31714 OTUs in the 140 samples. 


# 2. VENN

## soil management

```{r}
MyPalette <- c(forest = "#1167b1", meadow = "#fbc02d", organic = "#8a8a8a", conventional =  "#b71c1c")

# Note, relative is false, because I am using an alredy relative abundance transformed phyloseq object.
venn_no_prev <- ps_venn(
  ps_RA,
  "sample_type",
  fraction = 0,
  weight = FALSE,
  relative = FALSE,
  plot = TRUE,
  quantities = list(cex=0.7),
  fill = MyPalette
)
venn_no_prev

```


How many OTUs are shared by all sample types? I will pick the shared by all from the figure and divide it with the total OTU number to get the percentage

```{r}
shared_by_all <- 2570
total_OTUs <- 20610
  
shared_by_all/total_OTUs
```

12.5% of OTUs were shared by all management


## soil layer

```{r}
meta$new_depth <- meta$depth

meta$new_depth <- gsub("...", "-", meta$new_depth, fixed = TRUE)

meta$new_depth[meta$new_depth=="40-"] <- "40-80"

meta$new_depth[meta$new_depth=="0-10"] <- "0-10 cm"
meta$new_depth[meta$new_depth=="10-20"] <- "10-20 cm"
meta$new_depth[meta$new_depth=="20-30"] <- "20-30 cm"
meta$new_depth[meta$new_depth=="30-40"] <- "30-40 cm"
meta$new_depth[meta$new_depth=="40-80"] <- "40-80 cm"

sample_data(ps) <- sample_data(meta)

ps_RA <- transform(ps, "compositional")
  
# create your own color palette for sample types
MyPalette <- list(c('0-10 cm' = "#387212", '10-20 cm' = "#ADC476", '20-30 cm' = "#D8D2BA",'30-40 cm' = "#907852", '40-80 cm' = "#6A4C3A"))

venn_DEPTH <- ps_venn(
  ps_RA,
  "new_depth",
  fraction = 0,
  weight = FALSE,
  relative = TRUE,
  quantities = list(cex=0.7),
  plot = TRUE,
  fill = MyPalette[[1]]
)
venn_DEPTH

```


Again, how many OTUs are shared by all layers?

```{r}
shared_by_all <- 1007
total_OTUs <- 20610
  
shared_by_all/total_OTUs
```

4.9% of OTUs were shared by all layers


# OTU allocation to layers

How much of the OTUs in the dataset were found in the first, first two or first three soil layers?

## first three soil layers

```{r}
ps_x <- subset_samples(ps, depth!="30...40" & depth!="40...")

# You have only submitted samples, not OTUs. Define prevalence of each taxa
# (in how many samples did each taxa appear at least once)
prev0 = apply(X = otu_table(ps_x),
              MARGIN = ifelse(taxa_are_rows(ps_x), yes = 1, no = 2),
              FUN = function(x){sum(x > 0)})

# Execute prevalence filter, using `prune_taxa()` function
ps_x = prune_taxa((prev0 > 0), ps_x)
ps_x

```

In the first three layers: 20130 OTUs

```{r}
OTUs <- 20130
total_OTUs <- 20610
  
OTUs/total_OTUs
```

97.7% of all OTUs were found in the first 3 soil layers


How much of the OTUs in the dataset were found in the first, first two or first three soil layers?

## first two soil layers

```{r}
ps_x <- subset_samples(ps, depth!="20...30" & depth!="30...40" & depth!="40...")

# You have only submitted samples, not OTUs. Define prevalence of each taxa
# (in how many samples did each taxa appear at least once)
prev0 = apply(X = otu_table(ps_x),
              MARGIN = ifelse(taxa_are_rows(ps_x), yes = 1, no = 2),
              FUN = function(x){sum(x > 0)})

# Execute prevalence filter, using `prune_taxa()` function
ps_x = prune_taxa((prev0 > 0), ps_x)
ps_x

```

In the first three layers: 18392 OTUs

```{r}
OTUs <- 18392
total_OTUs <- 20610
  
OTUs/total_OTUs
```

89.2% of all OTUs were found in the first 2 soil layers


## first soil layer

```{r}
ps_x <- subset_samples(ps, depth=="0...10")

# You have only submitted samples, not OTUs. Define prevalence of each taxa
# (in how many samples did each taxa appear at least once)
prev0 = apply(X = otu_table(ps_x),
              MARGIN = ifelse(taxa_are_rows(ps_x), yes = 1, no = 2),
              FUN = function(x){sum(x > 0)})

# Execute prevalence filter, using `prune_taxa()` function
ps_x = prune_taxa((prev0 > 0), ps_x)
ps_x

```


```{r}
OTUs <- 14737
total_OTUs <- 20610
  
OTUs/total_OTUs
```

71.5% of all OTUs were found in the first soil layer




## combine the figures

```{r}
library(ggpubr)

fig <- ggarrange(venn_no_prev, venn_DEPTH,
                     ncol = 2, nrow = 1)

fig

```


















