---
title: "Untitled"
author: "Laura_Hakkinen"
date: "2024-06-04"
output: html_document
---

# RESULTS STEPS 8: Spearman correlations of depth with 5 most abundant phyla, classes, genera and trophic mode

**Note** all correlations are done without forest!

```{r}
library("ggpubr")
library('phyloseq')
library("cowplot")
library("dplyr")
library("ggplot2")
library("vegan")
library("microbiome")
library("tibble")
library(stringr)
library(reshape2)
library(tidyr)

setwd('\\\\ad.helsinki.fi\\home\\l\\lehakkin\\Desktop\\PROJECT_YONI_fungal_ITS')

load('ps_FINAL')
ps

load(file = 'ps_FG_with_NAs')#ps_FG
ps_FG

```


# Spearman rank correlation with depth without forest

## 1. Trophic modes

### 1.1. Remove forest and NAs

```{r}
# remove NAs from ps_FG

# this aggregation makes NAs into "Unknown"
ps_FG_x <- aggregate_rare(ps_FG, level = 'trophicMode', detection = 0/100, prevalence = 0/140)
ps_FG_x
# 8 taxa

# remove "Unknown"
allTaxa = taxa_names(ps_FG_x)
badTaxa = c("Unknown")
myTaxa <- allTaxa[!(allTaxa %in% badTaxa)]
ps_FG_x_pruned <- prune_taxa(myTaxa, ps_FG_x)
ps_FG_x_pruned
# 7 taxa

FG_nf <- subset_samples(ps_FG_x_pruned, sample_type != "forest")
FG_nf


FG_RA_nf <- transform(FG_nf, "compositional")
FG_RA_nf

df <- psmelt(FG_RA_nf)
```


### 1.2. Five most abundant Trophic modes

```{r}
df$year <- "2019"

# sum the values if same genus: Sum by Group Based on dplyr Package

x <- df %>%
  group_by(year, OTU) %>%
  summarise(mean = mean(Abundance, na.rm = TRUE), se = (sd(Abundance, na.rm = TRUE)/sqrt(length((Abundance)))))

y <- x %>%
    arrange(desc(mean)) %>%
    slice(1:5)
print(y)
```


### 1.3. Test for each Trophic mode separately

```{r}
taxa <- y$OTU

for (i in taxa) {
 df_x <- subset(df, OTU == i)
 print(i)
 x <- cor.test(df_x$Abundance, df_x$depth_numerical,  method = "spearman", na.rm = TRUE)
 print(x)
}

```



## 2. Phyla

### 2.1. Remove forest

```{r}
ps_nf <- subset_samples(ps, sample_type != "forest")
ps_nf


ps_RA_nf <- transform(ps_nf, "compositional")
ps_RA_nf

ps_RA_nf_phy <- aggregate_rare(ps_RA_nf, level = "phylum", detection = 0.0, prevalence = 0.0)
ps_RA_nf_phy

df <- psmelt(ps_RA_nf_phy)
```


### 2.2. Five most abundant phyla

```{r}
df$year <- "2019"

x <- df %>%
  group_by(year, OTU) %>%
  summarise(mean = mean(Abundance, na.rm = TRUE), se = (sd(Abundance, na.rm = TRUE)/sqrt(length((Abundance)))))
y <- x %>%
    arrange(desc(mean)) %>%
    slice(1:5)
print(y)
```

### 2.3. Test for each phyla separately


```{r}
taxa <- y$OTU

for (i in taxa) {
  df_x <- subset(df, OTU == i)
  print(i)
  x <- cor.test(df_x$Abundance, df_x$depth_numerical,  method = "spearman", na.rm = TRUE)
  print(x)
  }
```


## 3. Classes

### 3.1. Remove forest

```{r}
ps_nf <- subset_samples(ps, sample_type != "forest")
ps_nf


ps_RA_nf <- transform(ps_nf, "compositional")
ps_RA_nf

ps_RA_nf_cla <- aggregate_rare(ps_RA_nf, level = "class", detection = 0.0, prevalence = 0.0)
ps_RA_nf_cla

df <- psmelt(ps_RA_nf_cla)
```


### 3.2. Five most abundant phyla

```{r}
df$year <- "2019"

x <- df %>%
  group_by(year, OTU) %>%
  summarise(mean = mean(Abundance, na.rm = TRUE), se = (sd(Abundance, na.rm = TRUE)/sqrt(length((Abundance)))))
y <- x %>%
    arrange(desc(mean)) %>%
    slice(1:5)
print(y)
```

### 3.3. Test for each class separately


```{r}
taxa <- y$OTU

for (i in taxa) {
  df_x <- subset(df, OTU == i)
  print(i)
  x <- cor.test(df_x$Abundance, df_x$depth_numerical,  method = "spearman", na.rm = TRUE)
  print(x)
}

```


## 4. Genera

### 4.1. Remove forest

```{r}
ps_nf <- subset_samples(ps, sample_type != "forest")
ps_nf


ps_RA_nf <- transform(ps_nf, "compositional")
ps_RA_nf

ps_RA_nf_gen <- aggregate_rare(ps_RA_nf, level = "genus", detection = 0.0, prevalence = 0.0)
ps_RA_nf_gen

df <- psmelt(ps_RA_nf_gen)
```


### 4.2. Five most abundant genera

Note! Exclude the obes not classified at genus level!

```{r}
df$year <- "2019"

x <- df %>%
  group_by(year, OTU) %>%
  summarise(mean = mean(Abundance, na.rm = TRUE), se = (sd(Abundance, na.rm = TRUE)/sqrt(length((Abundance)))))
y <- x %>%
    arrange(desc(mean)) %>%
    slice(1:6)
print(y)
```


### 4.3. Test for each genera separately

We don't want to test for Leotiomycetes_unclassified 

```{r}
taxa <- y$OTU
taxa[! taxa %in% c("Leotiomycetes_unclassified")]

for (i in taxa) {
  df_x <- subset(df, OTU == i)
  print(i)
  x <- cor.test(df_x$Abundance, df_x$depth_numerical,  method = "spearman", na.rm = TRUE)
  print(x)
}
```



