---
title: "Untitled"
author: "Laura_Hakkinen"
date: "2024-06-19"
output: html_document
---

# RESULTS STEP 14: Spearman correlations with soil properties and fungal and AMF richness as well as with AMF relative abundance

Simple spearman rank correlation with richness and env. variables are done **WITHOUT** forest because forest soil is so different environment compared to meadow, organic and conventional soils

```{r}
library("ggpubr")
library('phyloseq')
library("cowplot")
library("dplyr")
library("ggplot2")
library("vegan")
library("microbiome")
library("tibble")
library(stringr)
library(reshape2)
library(tidyr)

setwd('\\\\ad.helsinki.fi\\home\\l\\lehakkin\\Desktop\\PROJECT_YONI_fungal_ITS')

load('ps_FINAL')
ps
```


# 1. remove forest

```{r}
ps_nf <- subset_samples(ps, sample_type!= "forest")

meta <- meta(ps_nf)
names(meta)
```


# 2. Spearman for richnesses


Soil properties to test against:

C_per_N
pH_H2O
C_g_per_kg
Feox_mmolkg
DOC_mgkg
N_gkg
depth_numerical
Porg_mgkg
log_root
TP_gkg
Pinorg_mgkg
Alox_mmolkg
PH2O_mgkg

## 2.1. Observed richness

```{r}
env <- c("C_per_N", "pH_H2O", "C_g_per_kg", "Feox_mmolkg", "DOC_mgkg", "N_gkg", "depth_numerical", "Porg_mgkg", "log_root", "TP_gkg", "Pinorg_mgkg", "Alox_mmolkg", "PH2O_mgkg")

for (i in env) {
   # Filter out rows with NA values in the columns of interest
  valid_data <- meta[!is.na(meta$observed) & !is.na(meta[[i]]), ]
  
  # Perform Spearman correlation test
  x <- cor.test(valid_data$observed, valid_data[[i]], method = "spearman")
  
  # Print the result
  print(paste("Correlation test for:", i))
  print(x)
  }


```


## 2.2. AMF_richness

```{r}
env <- c("C_per_N", "pH_H2O", "C_g_per_kg", "Feox_mmolkg", "DOC_mgkg", "N_gkg", "depth_numerical", "Porg_mgkg", "log_root", "TP_gkg", "Pinorg_mgkg", "Alox_mmolkg", "PH2O_mgkg")

for (i in env) {
   # Filter out rows with NA values in the columns of interest
  valid_data <- meta[!is.na(meta$AMF_richness) & !is.na(meta[[i]]), ]
  
  # Perform Spearman correlation test
  x <- cor.test(valid_data$AMF_richness, valid_data[[i]], method = "spearman")
  
  # Print the result
  print(paste("Correlation test for:", i))
  print(x)
  }

```


# 3. Spearman for AMF RA

```{r}
ps_nf_RA <- transform(ps_nf, "compositional")
ps_nf_RA

gm <- aggregate_rare(ps_nf_RA, level = "phylum", detection = 0, prevalence = 0 )
gm

df <- psmelt(gm)
df <- subset(df, OTU == "Glomeromycota")

env <- c("C_per_N", "pH_H2O", "C_g_per_kg", "Feox_mmolkg", "DOC_mgkg", "N_gkg", "depth_numerical", "Porg_mgkg", "log_root", "TP_gkg", "Pinorg_mgkg", "Alox_mmolkg", "PH2O_mgkg")

for (i in env) {
  # Perform Spearman correlation test
  x <- cor.test(df$Abundance, df[[i]], method = "spearman", na.rm = TRUE)
  
  # Print the result
  print(paste("Correlation test for:", i))
  print(x)
  }
```




