---
title: "Untitled"
author: "Laura_Hakkinen"
date: "2024-05-24"
output: html_document
---

# STEP 1: making phyloseq object

```{r}

library('phyloseq')
library("cowplot")
library("dplyr")
library("ggplot2")
library("vegan")
library("microbiome")
library("tibble")
library("readxl")
library(tidyverse)
```

# 1. import OTUs and taxa

```{r}
setwd('\\\\ad.helsinki.fi\\home\\l\\lehakkin\\Desktop\\YONI_ITS_analyses\\R_ITS_yoni\\Analyses_final\\RE_ANNOTATION_2024')

otu <- read.table("ITS_OTUs_mod.txt", sep="\t", header=FALSE)
tax <- read.table("ITS_OTUs.UNITEv10_sh_99.wang.taxonomy", sep="\t", header=FALSE)
```


More (4) observations (OTUS) in tax? check what is the deal. Get OTU names

```{r}
dif <- setdiff(tax$V1, otu$V1) 
dif
```

These OTUs are missing from OTU table but are present in tax table?
Let's remove them form tax table
Also, lets tweek the table row names and columns

```{r}
# modify otu table

colnames(otu)=otu[c(1),]
# erase the first row, as now it is doubled
otu=otu[-c(1),]

# make first column into rownames
rownames(otu) <- otu$`OTU ID`
otu <- otu[, -c(1)]
```


```{r}
# let's make a copy of tax table
tax.orig <- tax

# change column names
colnames(tax)[1] <- "OTU"
colnames(tax)[2] <- "taxa"

# and modify the tax, where in the str_remove_all . means any single character
tax <- tax %>%
  mutate(taxa = str_remove_all(taxa, ".__")) %>%
  separate(taxa,
           into = c("kingdom", "phylum", "class", "order", "family", "genus", "species"),
           sep = ";")

```

Everything went well but I got a warning message, because there are an extra ; at the end of the line so the last column is not made. which is correct. Let's check if we now have NAs, just to check everything is ok.

```{r}
sum(is.na(tax$OTU))
sum(is.na(tax$kingdom))
sum(is.na(tax$phylum))
sum(is.na(tax$class))
sum(is.na(tax$order))
sum(is.na(tax$family))
sum(is.na(tax$genus))
sum(is.na(tax$species))
```

Let's remove the parantheses and numbers

```{r}
tax$kingdom <- sub("\\(.*", "", tax$kingdom)
tax$phylum <- sub("\\(.*", "", tax$phylum)
tax$class <- sub("\\(.*", "", tax$class)
tax$order <- sub("\\(.*", "", tax$order)
tax$family <- sub("\\(.*", "", tax$family)
tax$genus <- sub("\\(.*", "", tax$genus)
tax$species <- sub("\\(.*", "", tax$species)
```


Check unique values of the higher taxons

```{r}
unique(tax$kingdom)
unique(tax$phylum)

```

## 1.2. remove the extra OTUs in tax table
"OTU57653" "OTU58305"

```{r}
tax <- subset(tax, OTU!="OTU57653")
tax <- subset(tax, OTU!="OTU58305")
```

# 2. import meta data

```{r}
samples <- read.csv2('\\\\ad.helsinki.fi\\home\\l\\lehakkin\\Desktop\\YONI_ITS_analyses\\R_ITS_yoni\\Analyses_final\\RE_ANNOTATION_2024\\sample_data_updated_root_biomass.csv')
# change some column names
colnames(samples)[which(names(samples) == "production_type")] <- "sample_type"

```

## 2.1. change otu table sample names

```{r}
x <- otu

x <- t(x)
x <- as.data.frame(x)
x <- rownames_to_column(x)

y <- samples[, c(1,2)]
x <- left_join(y, x, by = c("ID"="rowname"))
x <- x[, -c(1)]
rownames(x) <- x$sampleID
x <- x[, -c(1)]
x <- t(x)
otu <- x
rm(x)
rm(y)
library(dplyr)
otu <- as.data.frame(otu)
otu <- otu %>% mutate_if(is.character, as.numeric)
```



# 3. make phyloseq object

```{r}
rownames(tax) <- tax$OTU
tax <- tax[, -c(1)]
rownames(samples) <- samples$sampleID
samples <- samples[, -c(1)]
samples <- sample_data(samples)
otu <- as.matrix(otu)
otu <- otu_table(otu, taxa_are_rows = TRUE)
tax <- tax_table(as.matrix(tax))

ps <- phyloseq(otu, tax, samples)
ps

```


# remove unclassified phyla and controls

Fungi_unclassified
unclassified

```{r}
ps <- subset_samples(ps, sampleID != "0ctrl-1")
ps <- subset_samples(ps, sampleID != "0ctrl-2")
ps

# You have only submitted samples, not OTUs. Define prevalence of each taxa
# (in how many samples did each taxa appear at least once)
prev0 = apply(X = otu_table(ps),
              MARGIN = ifelse(taxa_are_rows(ps), yes = 1, no = 2),
              FUN = function(x){sum(x > 0)})

# Execute prevalence filter, using `prune_taxa()` function
ps = prune_taxa((prev0 > 0), ps)
ps
rm(prev0)
rm(tax.orig)

#I will remove two: "Fungi_unclassified" and "unclassified"
ps <- subset_taxa(ps, phylum != "Fungi_unclassified" & phylum != "unclassified")
ps

```

# 4. save phyloseq with sng

```{r}
ps_sng <- ps
setwd('\\\\ad.helsinki.fi\\home\\l\\lehakkin\\Desktop\\PROJECT_YONI_fungal_ITS')
save(ps_sng, file='ps_phyloseq_with_sng')
```


# 5. Remove sngs and save

```{r}
ps <- filter_taxa(ps, function (x) {sum(x > 0) > 1}, prune=TRUE)

setwd('\\\\ad.helsinki.fi\\home\\l\\lehakkin\\Desktop\\PROJECT_YONI_fungal_ITS')
save(ps, file='ps_FINAL')

ps
```



```{r}
summarize_phyloseq(ps)
```

# 6. Check low readcounts

```{r}
readcount(ps)
```

# 7. IMPORTANT! Bug in the TAX table!!

I noticed that there is a mistake in the taxonomy, so that sometimes I have "unclassified" annotation at lets say genus level, although I have annotation at lets say family level

In these cases I rather have at the genus level the annotation "family_unclassified"


```{r}
tax <- as.data.frame(tax_table(ps))
```


I have no unclassified at phylum level

change first at level:

class

```{r}
tax[tax$class == "unclassified", "phylum"]
```


```{r}
tax$class <- ifelse(tax$class == "unclassified", paste(tax$phylum, sep = "_", "unclassified"), as.character(tax$class))   
```


And same for all rest ranks, but I have to replace the "_unclassified_unclassified" with "_unclassified" afterwards


order

```{r}
tax$order <- ifelse(tax$order == "unclassified", paste(tax$class, sep = "_", "unclassified"), as.character(tax$order))   
```


```{r}

tax[] <- lapply(tax, gsub, pattern = "unclassified_unclassified", replacement = "unclassified", fixed = TRUE)

```


family


```{r}
tax$family <- ifelse(tax$family == "unclassified", paste(tax$order, sep = "_", "unclassified"), as.character(tax$family))   
```


```{r}

tax[] <- lapply(tax, gsub, pattern = "unclassified_unclassified", replacement = "unclassified", fixed = TRUE)

```


genus


```{r}
tax$genus <- ifelse(tax$genus == "unclassified", paste(tax$family, sep = "_", "unclassified"), as.character(tax$genus))   
```


```{r}

tax[] <- lapply(tax, gsub, pattern = "unclassified_unclassified", replacement = "unclassified", fixed = TRUE)

```


species


```{r}
tax$species <- ifelse(tax$species == "unclassified", paste(tax$genus, sep = "_", "unclassified"), as.character(tax$species))   
```


```{r}

tax[] <- lapply(tax, gsub, pattern = "unclassified_unclassified", replacement = "unclassified", fixed = TRUE)

```


# 8. Save

Save the new modified ps as the "final version"

```{r}
setwd('\\\\ad.helsinki.fi\\home\\l\\lehakkin\\Desktop\\PROJECT_YONI_fungal_ITS')

tax_table(ps) <- tax_table(as.matrix(tax))

save(ps, file='ps_FINAL')

ps
```





